\documentclass[14pt]{extarticle}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{enumitem}

\def\mathbi#1{\textbf{\em #1}}

\title{Elastic Net Regression}
\author{Nguyen Phu Nhan}
\date{}

\begin{document}

\maketitle

Elastic Net regression is a regularized version of linear regression that
combines the strengths of both Ridge regression and Lasso regression. Similar
to ordinary linear regression, its objective is to model the relationship
between a set of input variables and a continuous target variable. However,
Elastic Net introduces both an $L_1$ penalty and an $L_2$ penalty in order to
reduce model complexity, prevent overfitting, handle correlated features, and
perform feature selection.

Suppose we are given a dataset consisting of $n$ observations and $p$
features. Let

\[
\mathbf{X}
=
\begin{bmatrix}
x_{11} & x_{12} & \cdots & x_{1p} \\
x_{21} & x_{22} & \cdots & x_{2p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n1} & x_{n2} & \cdots & x_{np}
\end{bmatrix}
\]
be the feature matrix and

\[
\mathbi{y}
=
\begin{bmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{bmatrix}
\]
be the vector of target values.

As in linear regression, we absorb the bias term into the feature matrix by
appending a column of $1$'s to $\mathbf{X}$. The prediction function can then
be written as

\[
\hat{\mathbi{y}}
=
\mathbf{X}\mathbi{w},
\]
where $\mathbf{X} \in \mathbb{R}^{n\times p}$ is the feature matrix,
$\mathbi{w} \in \mathbb{R}^{p}$ is the coefficient vector, and
$\hat{\mathbi{y}} \in \mathbb{R}^{n}$ is the vector of predicted target
values.

In ordinary linear regression, we minimize the mean squared error loss

\[
L(\mathbi{w})
=
\frac{1}{2n}
\left\|
\mathbf{X}\mathbi{w}
-
\mathbi{y}
\right\|^2.
\]

Elastic Net regression modifies this objective function by adding both an
$L_1$ penalty and an $L_2$ penalty:

\[
L(\mathbi{w})
=
\frac{1}{2n}
\left\|
\mathbf{X}\mathbi{w}
-
\mathbi{y}
\right\|^2
+
\lambda_1
\left\|
\mathbi{w}
\right\|_1
+
\frac{\lambda_2}{2}
\left\|
\mathbi{w}
\right\|_2^2,
\]
where

\[
\left\|
\mathbi{w}
\right\|_1
=
\sum_{j=1}^{p}
|w_j|
\]
is the $L_1$ norm and

\[
\left\|
\mathbi{w}
\right\|_2^2
=
\sum_{j=1}^{p}
w_j^2
\]
is the squared $L_2$ norm.

Elastic Net is commonly parameterized using two hyperparameters:
$\alpha$ and $\text{l1\_ratio}$. We define

\[
\lambda_1
=
\alpha \cdot \text{l1\_ratio}
\]

and

\[
\lambda_2
=
\alpha \cdot
(1-\text{l1\_ratio}).
\]

Throughout the derivation, we use the shorthand notation
$\lambda_1$ and $\lambda_2$.

Equivalently,

\[
L(\mathbi{w})
=
\frac{1}{2n}
(\mathbf{X}\mathbi{w}-\mathbi{y})^\top
(\mathbf{X}\mathbi{w}-\mathbi{y})
+
\lambda_1
\sum_{j=1}^{p}
|w_j|
+
\frac{\lambda_2}{2}
\sum_{j=1}^{p}
w_j^2.
\]

The parameters $\lambda_1$ and $\lambda_2$ control the strength of
regularization.

\begin{itemize}[leftmargin=*]
\item If $\text{l1\_ratio}=1$, Elastic Net reduces to Lasso regression.
\item If $\text{l1\_ratio}=0$, Elastic Net reduces to Ridge regression.
\item If $\alpha=0$, the model reduces to ordinary linear regression.
\item Increasing $\alpha$ increases the overall strength of regularization.
\item Larger values of $\text{l1\_ratio}$ place more emphasis on feature
selection through the $L_1$ penalty.
\item Smaller values of $\text{l1\_ratio}$ place more emphasis on coefficient
shrinkage through the $L_2$ penalty.
\end{itemize}

The training objective is therefore

\[
\mathbi{w}^{\ast}
=
\operatorname*{argmin}_{\mathbi{w}}
\left[
\frac{1}{2n}
\left\|
\mathbf{X}\mathbi{w}
-
\mathbi{y}
\right\|^2
+
\lambda_1
\left\|
\mathbi{w}
\right\|_1
+
\frac{\lambda_2}{2}
\left\|
\mathbi{w}
\right\|_2^2
\right].
\]

Unlike ordinary linear regression and Ridge regression, there is generally no
closed-form solution because of the non-differentiability of the absolute
value function at zero. Consequently, Elastic Net is typically optimized using
coordinate descent.

For a coefficient $w_j$, define the partial residual

\[
r_i^{(j)}
=
y_i
-
\sum_{k\ne j}
x_{ik}w_k.
\]

Substituting this expression into the objective function yields

\[
L(w_j)
=
\frac{1}{2n}
\sum_{i=1}^{n}
\left(
r_i^{(j)}
-
x_{ij}w_j
\right)^2
+
\lambda_1|w_j|
+
\frac{\lambda_2}{2}w_j^2.
\]

Expanding the squared term gives

\[
L(w_j)
=
\frac{1}{2n}
\sum_{i=1}^{n}
\left[
(r_i^{(j)})^2
-
2x_{ij}r_i^{(j)}w_j
+
x_{ij}^2w_j^2
\right]
+
\lambda_1|w_j|
+
\frac{\lambda_2}{2}w_j^2.
\]

Define

\[
\rho_j
=
\sum_{i=1}^{n}
x_{ij}r_i^{(j)}
\]

and

\[
z_j
=
\sum_{i=1}^{n}
x_{ij}^2.
\]

Substituting these quantities into the objective function yields

\[
L(w_j)
=
\frac{z_j}{2n}w_j^2
-
\frac{\rho_j}{n}w_j
+
\lambda_1|w_j|
+
\frac{\lambda_2}{2}w_j^2
+
C,
\]

where

\[
C
=
\frac{1}{2n}
\sum_{i=1}^{n}
(r_i^{(j)})^2
\]

is a constant independent of $w_j$.

Ignoring constants, we minimize

\[
L(w_j)
=
\left(
\frac{z_j}{2n}
+
\frac{\lambda_2}{2}
\right)
w_j^2
-
\frac{\rho_j}{n}w_j
+
\lambda_1|w_j|.
\]

Because the absolute value function is piecewise-defined, we consider three
cases.

\paragraph{Case 1: $w_j > 0$}

In this case,

\[
|w_j|
=
w_j.
\]

The objective becomes

\[
L(w_j)
=
\left(
\frac{z_j}{2n}
+
\frac{\lambda_2}{2}
\right)
w_j^2
-
\frac{\rho_j}{n}w_j
+
\lambda_1 w_j.
\]

Differentiating with respect to $w_j$ gives

\[
\frac{dL}{dw_j}
=
\left(
\frac{z_j}{n}
+
\lambda_2
\right)
w_j
-
\frac{\rho_j}{n}
+
\lambda_1.
\]

Setting the derivative equal to zero,

\[
\left(
z_j
+
n\lambda_2
\right)
w_j
=
\rho_j
-
n\lambda_1.
\]

Therefore,

\[
w_j
=
\frac{\rho_j-n\lambda_1}
     {z_j+n\lambda_2}.
\]

This solution is valid only when

\[
\rho_j
>
n\lambda_1.
\]

\paragraph{Case 2: $w_j < 0$}

In this case,

\[
|w_j|
=
-w_j.
\]

The objective function becomes

\[
L(w_j)
=
\left(
\frac{z_j}{2n}
+
\frac{\lambda_2}{2}
\right)
w_j^2
-
\frac{\rho_j}{n}w_j
-
\lambda_1 w_j.
\]

Differentiating,

\[
\frac{dL}{dw_j}
=
\left(
\frac{z_j}{n}
+
\lambda_2
\right)
w_j
-
\frac{\rho_j}{n}
-
\lambda_1.
\]

Setting the derivative equal to zero,

\[
\left(
z_j
+
n\lambda_2
\right)
w_j
=
\rho_j
+
n\lambda_1.
\]

Hence,

\[
w_j
=
\frac{\rho_j+n\lambda_1}
     {z_j+n\lambda_2}.
\]

This solution is valid only when

\[
\rho_j
<
-n\lambda_1.
\]

\paragraph{Case 3: $w_j = 0$}

If

\[
-n\lambda_1
\le
\rho_j
\le
n\lambda_1,
\]

the minimum occurs at

\[
w_j
=
0.
\]

Combining all three cases, we obtain

\[
w_j
=
\begin{cases}
\dfrac{\rho_j-n\lambda_1}
      {z_j+n\lambda_2},
&
\rho_j > n\lambda_1,
\\[12pt]
0,
&
|\rho_j|\le n\lambda_1,
\\[12pt]
\dfrac{\rho_j+n\lambda_1}
      {z_j+n\lambda_2},
&
\rho_j < -n\lambda_1.
\end{cases}
\]

This piecewise expression can be written compactly using the
soft-thresholding operator

\[
S(\rho,\lambda)
=
\begin{cases}
\rho-\lambda,
&
\rho>\lambda,
\\[6pt]
0,
&
|\rho|\le\lambda,
\\[6pt]
\rho+\lambda,
&
\rho<-\lambda.
\end{cases}
\]

Substituting

\[
\lambda_1
=
\alpha \cdot \text{l1\_ratio}
\]

and

\[
\lambda_2
=
\alpha \cdot
(1-\text{l1\_ratio}),
\]

the coordinate descent update becomes

\[
w_j
=
\frac{
S\!\left(
\rho_j,
n\alpha \cdot \text{l1\_ratio}
\right)
}
{
z_j
+
n\alpha
\left(
1-\text{l1\_ratio}
\right)
}.
\]

The soft-thresholding operator shrinks coefficients toward zero and sets small
coefficients exactly equal to zero, while the Ridge component further shrinks
all coefficients through the denominator.

Consequently, Elastic Net performs both feature selection and coefficient
shrinkage simultaneously. This makes it particularly useful when the number of
features is large, when many features are irrelevant, or when predictors are
highly correlated.

\end{document}
