\documentclass[14pt]{extarticle}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{enumitem}

\def\mathbi#1{\textbf{\em #1}}

\title{Lasso Regression}
\author{Nguyen Phu Nhan}
\date{}

\begin{document}

\maketitle

Lasso (Least Absolute Shrinkage and Selection Operator) regression is a
regularized version of linear regression. Similar to ordinary linear
regression, its objective is to model the relationship between a set of input
variables and a continuous target variable. However, Lasso regression adds a
penalty term to the loss function in order to reduce model complexity, prevent
overfitting, and perform feature selection.

Suppose we are given a dataset consisting of $n$ observations and $p$
features. Let

\[
\mathbf{X} =
\begin{bmatrix}
x_{11} & x_{12} & \cdots & x_{1p} \\
x_{21} & x_{22} & \cdots & x_{2p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n1} & x_{n2} & \cdots & x_{np}
\end{bmatrix}
\]
be the feature matrix and

\[
\mathbi{y}
=
\begin{bmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{bmatrix}
\]
be the vector of target values.

As in linear regression, we absorb the bias term into the feature matrix by
appending a column of $1$'s to $\mathbf{X}$. The prediction function can then
be written as

\[
\hat{\mathbi{y}}
=
\mathbf{X}\mathbi{w},
\]
where $\mathbf{X} \in \mathbb{R}^{n \times p}$ is the feature matrix,
$\mathbi{w} \in \mathbb{R}^{p}$ is the coefficient vector, and
$\hat{\mathbi{y}} \in \mathbb{R}^{n}$ is the vector of predicted target
values.

In ordinary linear regression, we minimize the mean squared error loss

\[
L(\mathbi{w})
=
\frac{1}{2n}
\left\|
\mathbf{X}\mathbi{w}
-
\mathbi{y}
\right\|^2.
\]

Lasso regression modifies this objective function by adding an $L_1$
regularization term:

\[
L(\mathbi{w})
=
\frac{1}{2n}
\left\|
\mathbf{X}\mathbi{w}
-
\mathbi{y}
\right\|^2
+
\lambda
\left\|
\mathbi{w}
\right\|_1,
\]
where

\[
\left\|
\mathbi{w}
\right\|_1
=
\sum_{j=1}^{p}
|w_j|
\]
is the $L_1$ norm of the coefficient vector, and $\lambda \ge 0$ is the
regularization parameter.

Equivalently,

\[
L(\mathbi{w})
=
\frac{1}{2n}
(\mathbf{X}\mathbi{w}-\mathbi{y})^\top
(\mathbf{X}\mathbi{w}-\mathbi{y})
+
\lambda
\sum_{j=1}^{p}
|w_j|.
\]

The parameter $\lambda$ controls the strength of regularization:
\begin{itemize}[leftmargin=*]
\item If $\lambda = 0$, the model reduces to ordinary linear regression.
\item As $\lambda$ increases, larger penalties are imposed on the coefficients.
\item Very large values of $\lambda$ force many coefficients toward zero.
\end{itemize}

The training objective is therefore

\[
\mathbi{w}^\ast
=
\operatorname*{argmin}_{\mathbi{w}}
\left[
\frac{1}{2n}
\left\|
\mathbf{X}\mathbi{w}
-
\mathbi{y}
\right\|^2
+
\lambda
\left\|
\mathbi{w}
\right\|_1
\right].
\]

Unlike ordinary linear regression, there is generally no closed-form solution
for this optimization problem because the absolute value function is not
differentiable at zero.

For this reason, iterative optimization algorithms are used. One of the most
popular approaches is \emph{coordinate descent}, which updates one coefficient
at a time while keeping all other coefficients fixed.

For a coefficient $w_j$, define the partial residual

\[
r_i^{(j)}
=
y_i
-
\sum_{k\ne j}
x_{ik}w_k.
\]

Substituting this expression into the objective function yields

\[
L(w_j)
=
\frac{1}{2n}
\sum_{i=1}^{n}
\left(
r_i^{(j)}
-
x_{ij}w_j
\right)^2
+
\lambda |w_j|.
\]

Let

\[
\rho_j
=
\sum_{i=1}^{n}
x_{ij}r_i^{(j)}
\]

and

\[
z_j
=
\sum_{i=1}^{n}
x_{ij}^2.
\]

The optimal update for coefficient $w_j$ can be derived by minimizing the
one-dimensional objective function

\[
L(w_j)
=
\frac{1}{2n}
\sum_{i=1}^{n}
\left(
r_i^{(j)}
-
x_{ij}w_j
\right)^2
+
\lambda |w_j|.
\]

Expanding the squared term gives

\[
L(w_j)
=
\frac{1}{2n}
\sum_{i=1}^{n}
\left[
(r_i^{(j)})^2
-
2x_{ij}r_i^{(j)}w_j
+
x_{ij}^2w_j^2
\right]
+
\lambda |w_j|.
\]

Define

\[
\rho_j
=
\sum_{i=1}^{n}
x_{ij}r_i^{(j)}
\]

and

\[
z_j
=
\sum_{i=1}^{n}
x_{ij}^2.
\]

Substituting these quantities into the objective function yields

\[
L(w_j)
=
\frac{z_j}{2n}w_j^2
-
\frac{\rho_j}{n}w_j
+
\lambda |w_j|
+
C,
\]

where

\[
C
=
\frac{1}{2n}
\sum_{i=1}^{n}
(r_i^{(j)})^2
\]

is a constant independent of $w_j$. Since constants do not affect the
location of the minimum, we only need to minimize

\[
L(w_j)
=
\frac{z_j}{2n}w_j^2
-
\frac{\rho_j}{n}w_j
+
\lambda |w_j|.
\]

Because the absolute value function is piecewise-defined, we consider three
cases.

\paragraph{Case 1: $w_j > 0$}

In this case,

\[
|w_j|
=
w_j.
\]

Therefore,

\[
L(w_j)
=
\frac{z_j}{2n}w_j^2
-
\frac{\rho_j}{n}w_j
+
\lambda w_j.
\]

Taking the derivative with respect to $w_j$ gives

\[
\frac{dL}{dw_j}
=
\frac{z_j}{n}w_j
-
\frac{\rho_j}{n}
+
\lambda.
\]

Setting the derivative equal to zero,

\[
\frac{z_j}{n}w_j
-
\frac{\rho_j}{n}
+
\lambda
=
0.
\]

Multiplying both sides by $n$,

\[
z_j w_j
-
\rho_j
+
n\lambda
=
0.
\]

Hence,

\[
w_j
=
\frac{\rho_j-n\lambda}{z_j}.
\]

This solution is valid only when $w_j>0$, which implies

\[
\rho_j
>
n\lambda.
\]

\paragraph{Case 2: $w_j < 0$}

In this case,

\[
|w_j|
=
-w_j.
\]

The objective function becomes

\[
L(w_j)
=
\frac{z_j}{2n}w_j^2
-
\frac{\rho_j}{n}w_j
-
\lambda w_j.
\]

Differentiating,

\[
\frac{dL}{dw_j}
=
\frac{z_j}{n}w_j
-
\frac{\rho_j}{n}
-
\lambda.
\]

Setting the derivative equal to zero,

\[
\frac{z_j}{n}w_j
-
\frac{\rho_j}{n}
-
\lambda
=
0.
\]

Multiplying both sides by $n$,

\[
z_j w_j
-
\rho_j
-
n\lambda
=
0.
\]

Therefore,

\[
w_j
=
\frac{\rho_j+n\lambda}{z_j}.
\]

This solution is valid only when $w_j<0$, which implies

\[
\rho_j
<
-n\lambda.
\]

\paragraph{Case 3: $w_j = 0$}

If

\[
-n\lambda
\le
\rho_j
\le
n\lambda,
\]

neither of the previous solutions satisfies its sign constraint. In this
region, the minimum occurs at

\[
w_j
=
0.
\]

Combining all three cases, we obtain

\[
w_j
=
\begin{cases}
\dfrac{\rho_j-n\lambda}{z_j},
&
\rho_j > n\lambda,
\\[10pt]
0,
&
|\rho_j|\le n\lambda,
\\[10pt]
\dfrac{\rho_j+n\lambda}{z_j},
&
\rho_j < -n\lambda.
\end{cases}
\]

This piecewise expression can be written compactly using the
soft-thresholding operator

\[
S(\rho,\lambda)
=
\begin{cases}
\rho-\lambda,
&
\rho>\lambda,
\\[6pt]
0,
&
|\rho|\le\lambda,
\\[6pt]
\rho+\lambda,
&
\rho<-\lambda.
\end{cases}
\]

Thus, the coordinate descent update becomes

\[
w_j
=
\frac{S(\rho_j,n\lambda)}
{z_j}.
\]

The soft-thresholding operator shrinks coefficients toward zero and sets
small coefficients exactly equal to zero. Consequently, Lasso regression
performs feature selection automatically while fitting the model.

Consequently, Lasso regression performs both coefficient estimation and feature
selection simultaneously, making it particularly useful when the number of
features is large or when many features are irrelevant.

\end{document}